---
title: "drcTMB"
author: "Daniel Gerhard"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{drcTMB}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

The package drcTMB provides functions to fit dose-response curves using the Template Model Builder (TMB) framework.

## The drmTMB function

The main function in the drcTMB package is `drmTMB()`. There is some similarity to the `drm()` function from the drc package, with a simplified input of response and dose variables through a formula interface, and specifying a nonlinear model through the `model` argument. Available models include "logistic", "loglogistic", "weibull1", "weibull2", and "lognormal".

Each of these models has the same parameter names, b1 and b2 for the asymptotes, b3 for the steepness, b4 for the location of the inflection point on the dose scale, and b5 for an optional asymmetry parameter. The asymmetry parameter b5 is estimated on a logarithmic scale.
The `start` argument provides starting values for the parameters. The `fix` argument (TRUE/FALSE) allows to fix parameters at these starting values.

```{r}
library(drcTMB)
library(drcData)
data(ryegrass)
# Fitting a 3-parameter log-logistic model
mod <- drmTMB(rootl ~ conc, data=ryegrass, model="loglogistic", 
              start=c(8, 0, 1, 2, 0), fix=c(FALSE, TRUE, FALSE, FALSE, TRUE))
print(mod)
```

With the `predict()` method, we can visualize the fitted curve.
```{r}
newdata <- data.frame(conc = exp(seq(-0.5, log(30), length.out = 100)))
newdata$pred <- predict(mod, newdata = newdata)
library(ggplot2)
ggplot(ryegrass, aes(x = log(conc), y = rootl)) +
  geom_point() +
  geom_line(data = newdata, aes(x = log(conc), y = pred), color = "blue") +
  labs(x = "Log Concentration", y = "Root Length") +
  theme_bw()
```

## Effective dose estimation

The `ed()` function can be used to estimate effective doses (e.g., ED50, ED90) from the fitted model, given a specified response level (`respLev`).
```{r}
(effdose <- ed(mod, respLev = c(25, 50, 75)))
```

Simultaneous confidence intervals can be constructed through the package multcomp.
```{r, message=FALSE, warning=FALSE}
library(multcomp)
linfct <- diag(3)
rownames(linfct) <- c("ED25", "ED50", "ED75")
effdose_glht <- glht(effdose, linfct = linfct)
confint(effdose_glht)
```

## Non-normal response distributions

The `family` argument allows to specify the distribution of the response as "gaussian", "poisson", "binomial", or "beta".
To improve optimisation, the `link` argument allows us to estimate the asymptote parameters b1 and b2 on a transformed scale. Furthermore, we can add constraints on the parameters through the `lower` and `upper` arguments with vectors for the lower and upper bounds of the parameters, respectively.
```{r}
data(daphnids)
d48 <- subset(daphnids, time == "48h")

dmod <- drmTMB(cbind(no, total-no) ~ dose, data=d48, model="loglogistic", family="binomial", 
               start=c(0, 1, 1, 1000, 0), fix=c(TRUE, TRUE, FALSE, FALSE, FALSE))
dmod
```

```{r}
ndata <- data.frame(dose = exp(seq(3, log(22000), length.out = 100)))
ndata$pred <- predict(dmod, newdata = ndata)

ggplot(d48, aes(x = dose, y = no/total)) +
  geom_point() +
  geom_line(data = ndata, aes(x = dose, y = pred), color = "blue") +
  labs(x = "Dose", y = "Proportion of Immobile Daphnids") +
  theme_bw() +
  scale_x_continuous(breaks = c(100, 500, 2500, 10000), trans="log")
```


## Additional parameters

We can include a right-hand side formula for each of the model parameters through the argument `fform`, adding covariates on the parameter level.

```{r}
# using the full daphnids dataset with groups 24h and 48h
dmodg <- drmTMB(cbind(no, total-no) ~ dose, data=daphnids, 
                model="loglogistic", family="binomial",
                fform=list(X1=~1, X2=~1, X3=~0+time, X4=~0+time, X5=~1),
                start=c(0, 1, 1, 1, 1000, 1000, 0), 
                fix=c(TRUE, TRUE, FALSE, FALSE, FALSE, FALSE, FALSE))
dmodg
```

```{r}
ndatag <- expand.grid(dose = exp(seq(3, log(22000), length.out = 100)), 
                      time = c("24h", "48h"))
ndatag$pred <- predict(dmodg, newdata = ndatag)

ggplot(daphnids, aes(x = dose, y = no/total, colour=time)) +
  geom_point() +
  geom_line(data = ndatag, aes(x = dose, y = pred, colour=time)) +
  labs(x = "Dose", y = "Proportion of Immobile Daphnids") +
  theme_bw() +
  scale_x_continuous(breaks = c(100, 500, 2500, 10000), trans="log")
```


### Estimating effective doses for a linear function of parameters
```{r, message=FALSE, warning=FALSE}
linfct <- list("24h"=rbind(c(1, 0, 0, 0, 0, 0, 0),
                           c(0, 1, 0, 0, 0, 0, 0),
                           c(0, 0, 1, 0, 0, 0, 0),
                           c(0, 0, 0, 0, 1, 0, 0),
                           c(0, 0, 0, 0, 0, 0, 1)),
               "48h"=rbind(c(1, 0, 0, 0, 0, 0, 0),
                           c(0, 1, 0, 0, 0, 0, 0),
                           c(0, 0, 0, 1, 0, 0, 0),
                           c(0, 0, 0, 0, 0, 1, 0),
                           c(0, 0, 0, 0, 0, 0, 1)))
(ed75 <- ed(dmodg, respLev = 75, linfct = linfct))

lfdiff <- rbind("24h"=c(1, 0),
                "48h"=c(0, 1),
                "48h-24h"=c(-1, 1))
edose_glht <- glht(ed75, linfct = lfdiff)
confint(edose_glht)
```

## Random effects

Fitting a 5-parameter log-logistic model to the vinclozolin dataset, which contains luminescence measurements of a dose-response experiment with 6 replications. The model includes random effects for the lower and upper asymptote b1 and b2, and the location of the inflection point b4, with a random intercept for each replication.
Given that there is not much information about the upper asymptote available in the sample, there is some dependency on the starting values. But the model can still be useful as an example for fitting a nonlinear mixed-effect model with the drmTMB function. The `lower` argument is used to constrain the fixed effect parameters to be positive, to improve optimisation.

```{r, message=FALSE, warning=FALSE}
data(vinclozolin)
vmod <- drmTMB(effect ~ conc, data=vinclozolin, 
               model="loglogistic", 
               rform=b1 + b2 + b4 ~ 1 | exper,
               start = c(1700, 100, 0.5, 0.01, 0), 
               fix = c(FALSE, FALSE, FALSE, FALSE, FALSE), 
               lower = c(0, 0, -Inf, 0, 0))
```

We can visualise the predictions conditional on the random effects.

```{r, message=FALSE, warning=FALSE}
ndatav <- expand.grid(conc = exp(seq(-5, log(6), length.out = 500)), 
                      exper = levels(vinclozolin$exper))

# predict.drmTMB by default provides predictions conditional on the random effects
ndatav$pred <- predict(vmod, newdata = ndatav)
# prediction conditional on random effects = 0
ndatav$predpop <- predict(vmod, newdata = ndatav, random=FALSE)

ggplot(vinclozolin, aes(x = log(conc), y = effect, colour=exper)) +
  geom_point() +
  geom_line(data = ndatav, aes(x = log(conc), y = pred, colour=exper)) +
  geom_line(data = ndatav, aes(x = log(conc), y = predpop), 
            colour="black", linewidth=1.2) +
  labs(x = "log Concentration", y = "Fluorescence") +
  theme_bw() 

```


### Estimating effective doses in a mixed-effect model

The `ed()` function is using the delta method to estimates effective doses and their corresponding standard errors. The standard error estimation includes by default the estimated variance components of the random effects.

```{r, message=FALSE, warning=FALSE}
(edv <- ed(vmod, respLev=50, random=TRUE))
```

We can additionally include the residual variance in the standard error estimation.
```{r, message=FALSE, warning=FALSE}
(edvr <- ed(vmod, respLev=50, random=TRUE, residual=TRUE))
```
