---
title: "drcTMB"
author: "Daniel Gerhard"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{drcTMB}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

The package drcTMB provides functions to fit dose-response curves using the Template Model Builder (TMB) framework.

## The drmTMB function

The main function in the drcTMB package is `drmTMB()`. There is some similarity to the `drm()` function from the drc package, with a simplified input of response and dose variables through a formula interface, and specifying a nonlinear model through the `model` argument. Available models include "logistic", "loglogistic", "weibull1", "weibull2", and "lognormal".

Each of these models has the same parameter names, b1 and b2 for the asymptotes, b3 for the steepness, b4 for the location of the inflection point on the dose scale, and b5 for an optional asymmetry parameter. The asymmetry parameter b5 is estimated on a logarithmic scale.
The `start` argument provides starting values for the parameters. The `fix` argument (TRUE/FALSE) allows to fix parameters at these starting values.

```{r}
library(drcTMB)
library(drcData)
data(ryegrass)
# Fitting a 3-parameter log-logistic model
mod <- drmTMB(rootl ~ conc, data=ryegrass, model="loglogistic", 
              start=c(8, 0, 1, 2, 0), fix=c(FALSE, TRUE, FALSE, FALSE, TRUE))
print(mod)
```

With the `predict()` method, we can visualize the fitted curve.
```{r}
newdata <- data.frame(conc = exp(seq(-0.5, log(30), length.out = 100)))
newdata$pred <- predict(mod, newdata = newdata)
library(ggplot2)
ggplot(ryegrass, aes(x = log(conc), y = rootl)) +
  geom_point() +
  geom_line(data = newdata, aes(x = log(conc), y = pred), color = "blue") +
  labs(x = "Log Concentration", y = "Root Length") +
  theme_bw()
```

## Effective dose estimation

The `ed()` function can be used to estimate effective doses (e.g., ED50, ED90) from the fitted model, given a specified response level (`respLev`).
```{r}
(effdose <- ed(mod, respLev = c(25, 50, 75)))
```

Simultaneous confidence intervals can be constructed through the package multcomp.
```{r, message=FALSE, warning=FALSE}
library(multcomp)
linfct <- diag(3)
rownames(linfct) <- c("ED25", "ED50", "ED75")
effdose_glht <- glht(effdose, linfct = linfct)
confint(effdose_glht)
```

## Non-normal response distributions

The `family` argument allows to specify the distribution of the response as "gaussian", "poisson", "binomial", or "beta".
To improve optimisation, the `link` argument allows us to estimate the asymptote parameters b1 and b2 on a transformed scale. Furthermore, we can add constraints on the parameters through the `lower` and `upper` arguments with vectors for the lower and upper bounds of the parameters, respectively.
```{r}
data(daphnids)
d48 <- subset(daphnids, time == "48h")

dmod <- drmTMB(cbind(no, total-no) ~ dose, data=d48, model="loglogistic", family="binomial", 
               start=c(0, 1, 1, 1000, 0), fix=c(TRUE, TRUE, FALSE, FALSE, FALSE))
dmod
```

```{r}
ndata <- data.frame(dose = exp(seq(3, log(22000), length.out = 100)))
ndata$pred <- predict(dmod, newdata = ndata)

ggplot(d48, aes(x = log(dose), y = no/total)) +
  geom_point() +
  geom_line(data = ndata, aes(x = log(dose), y = pred), color = "blue") +
  labs(x = "Log Dose", y = "Proportion of Immobile Daphnids") +
  theme_bw()
```